Evaluating the Performance and Energy Efficiency of the \textsc{COSMO-ART} model system

### 2014-04-25 trunk/paper_enahpc14/Sec3:Environment-setup.tex
Manuel's changes

Benchmarks presented  in this work  were performed on  three different
platforms  and can  be reproduced  within an  expected  variance, thus
allowing a  fair comparison between the baseline  and future milestone
versions of COSMO-ART.

\begin{itemize}
\item \textsc{Monch (CSCS)} is a 10 rack NEC-provided cluster composed
  of 312 compute  nodes and utilised by members  of the Swiss Platform
  for Advanced Scientific  Computing (PASC) project \citep{PASC}. Each
  node comprises  two Intel Ivy Bridge EP  E5-2660 ten-core processors
  operating at  2.2 GHz connected  via InfiniBand Mellanox  SX6036 and
  FDR switches (56 Gb/s speed), and is equipped with 32GB of DDR3 1600
  MHz  RAM.   For our  experiments,  a  full  rack constituted  of  52
  standard compute  nodes was used.  \textsc{Monch} is  slated to stay
  in  service  without  hardware  upgrade  for  the  duration  of  the
  Exa2Green  project,  allowing   an  identical  configuration  to  be
  recreated for future assessments of the baseline.\\

\item \textsc{Pilatus (CSCS)} is  a dual-socket eight-core Intel Sandy
  Bridge EP based cluster composed of 42 compute nodes and used as Piz
  Daint pre-post  processing cluster at  CSCS.  The 2 login  nodes and
  the 42 computes nodes consists in 11 twin-pair Intel E5-Series DALCO
  r2264i4t 2U scalable compute modules. Each module contains 4 compute
  nodes based  on two Intel  Xeon E5-2670 processors operating  at 2.6
  GHz connected  via InfiniBand Mellanox  SX6036 and FDR  switches (56
  Gb/s speed), and is equipped with 64GB of DDR3 1600 MHz RAM. For our
  experiments, a  full rack constituted  of 42 standard  compute nodes
  was  used. \textsc{Pilatus}  is based  on  Intel's second-generation
  processors, conventional in  HPC systems and known to  be more power
  consuming than Ivy Bridge successor.\\

\item \textsc{Tintorrum (UJI)} is  a heterogeneous cluster composed of
  28 compute nodes.  For our  experiments only a subset of these nodes
  were considered.   This set is composed  of 16 nodes,  each of which
  includes  two Intel  Xeon  E5645 hexa-core  processors  at 2.40  GHz
  connected via  Infiniband QDR (Mellanox MTS3600  switch).  Each core
  has  32 KB  instruction  and 32  KB data  L1  caches and  256 KB  L2
  cache. The 6 cores  share a 12 MB L3 cache and  the platform has 288
  GB  of DDR3 1333  MHz.  A  full tracing  experiment is  conducted on
  \textsc{Tintorrum}, to  capture an overall  power profile at  a much
  finer resolution, and to get more insights into the MPI blocking and
  polling influences on power savings.
\end{itemize}

\subsection{Power-performance measurement framework}
\label{subsec:3.3}

We present  here two different  frameworks deployed on HPC  systems to
measure  the  power  consumption   and  performance  of  the  baseline
execution.

\subsubsection{CSCS - E3METER metering products}

Supercomputer  clusters considered  for our  experiments at  the Swiss
National Supercomputing Center (CSCS)  of ETH Zurich are equipped with
E3METER Intelligent  Power Strips (IPS)  and Monitors (IPM)  which are
high   quality   electricity  meters   released   by  Riedo   Networks
\citep{Riedonetworks},   that  enable   to  monitor   and   log  power
consumption  of the IT  infrastructure as  well as  constantly analyze
line  voltage,  current, power-factor,  frequency  with 1\%  accuracy.
Using  reliable narrowband  powerline communication  (PLC) technology,
all  metering and  power  quality  data from  each  IPS are  centrally
collected  by the E3METER  Data Concentrator,  via the  existing power
cables thus  avoiding the need for  extra cabling.  This  data is made
available via  SNMP, HTTP, TELNET  through the built-in  Fast Ethernet
port.   Time  synchronisation  is  guaranteed by  using  NTP  servers.
Measured  data is  accessed  through the  open  source Cacti  software
including the E3METER Cacti Plugin  to scan the entire PLC network and
monitor in real-time the power usage of individual rack, recorded in 5
minutes interval periods.

\subsubsection{UHAM - UJI}

For our experiments at Jaume I University (UJI), we employed a version
of the integrated framework presented in~\cite{energy13} that works in
combination with  Extrae and Paraver, which  are profiling/tracing and
visualization tools, respectively.

%The left part of the \vref{fig:Lustre} offers a graphical representation of
%the Lustre architecture; the right depicts the tracing and profiling framework.
COSMO-ART  is  compiled  using  the Extrae  compiler  wrappers,  which
automatically  instrument  the  Fortran   code  of  the  model.  Next,
COSMO-ART  is run  on the  nodes, thus  dissipating certain  amount of
power.  These  nodes are connected  to power measurement  devices that
account for  the dissipated power/consumed  energy and send  the power
data to  the tracing server.  The client,  meanwhile, sends start/stop
primitives in  order to gather  captured data by the  watt-meters onto
the tracing server, where an instance of the \pmlib server is running.

Once  the run  is finished,  a file  containing the  power  profile is
created using the power data  received from the tracing server and the
instrumentation post-processing generates the performance trace files.
All  these files  are  combined then  in  Paraver which  allows us  to
visualise the performance trace and the power profile of COSMO-ART all
together.

%Once COSMO-ART run is finished, the VampirTrace \pmlib plugin receives
%the  power   data  from  the  tracing   server.   The  instrumentation
%post-processing generates  the performance trace files  and the \pmlib
%plugin inserts the power data into them.

%In  addition  to the  power  measurements,  we  also account  for  the
%resource utilization values  of the nodes: CPU load,  memory usage and
%storage device utilization. % and network utilization.

%We  run special  \pmlib  server  instances on  the  server nodes  that
%retrieve these  values from the \texttt{proc}  file system (leveraging
%the  \texttt{psutil} Python library).   Thus, \pmlib  plugin instances
%running  with the  instrumented  application connect  with the  \pmlib
%servers.   Finally,   using  the   Vampir   visualization  tool,   the
%power-performance traces  can be easily  analyzed through a  series of
%plots and statistics.

@@

The COSMO-ART baseline  is a pure MPI based  Fortran~90 code currently
running on distributed multi-core systems only.  The software stack on
both CSCS  platforms was controlled using the  modules framework which
gives an easy and flexible mechanism  to access to all of the provided
compilers, tools  and applications.  For our  initial benchmarking, we
opted for the GNU  compiler (gcc/4.8.1 on \textsc{Monch}, gcc/4.8.2 on
\textsc{Pilatus}) using  the -O3 compiler  flag in favor of  the intel
compiler (14.0.1),  delivering inferior performance.   In addition, we
installed the  MPICH2 implementation of MPI (mvapich2/1.9)  as well as
the commonly used  HDF5 (1.8.12) and NetCDF (4.3.1)  libraries for the
management  of  extremely large  and  complex  data collections.   All
computes nodes  have an operating system based  on GNU/Linux featuring
``2.6.32-358.11.1.el6.x86\_64''    kernel   in    \textsc{Monch}   and
``3.0.101-0.15-default'' kernel in \textsc{Pilatus}.

A snapshot of the code,  which includes at least conceptually, all the
information needed  to reproduce the  energy-to-solution benchmarks of
COSMO-ART, was produced and run on a 1040 cores using 20~MPI tasks per
node on \textsc{Monch} and on a 1344 cores using 16~MPI tasks per node
on  \textsc{Pilatus}.   The  calculated   region  was  mapped  to  the
participating  processors  using   a  2D-partitioning  strategy.   The
distribution along the $x$ and $y$ coordinates was defined by setting:
$nprocx=40$  and $nprocy=26$  for \textsc{Monch}  and  $nprocx=28$ and
$nprocy=24$  as  $nprocx$  is   usually  kept  bigger  than  $nprocy$.
Besides,  as this  version doesn't  make use  of the  traditional GRIB
library, we  specified $nprocio=0$  for GRIB I/O.   Hyperthreading was
not considered in this study  as it previously revealed that it always
led to higher energy-to-solution.

>>>>

Benchmarks presented  in this work  were performed on  three different
platforms  and can  be reproduced  within an  expected  variance, thus
allowing a  fair comparison between the baseline  and future milestone
versions of \textsc{Cosmo-art}.

\begin{itemize}
\item \textsc{Monch} is a 10 rack NEC-provided cluster composed
  of 312 compute  nodes and utilized by members  of the Swiss Platform
  for Advanced Scientific  Computing (PASC) project~\citep{PASC} at CSCS. Each
  node comprises two Intel Ivy Bridge EP  E5-2660 ten-core processors
  operating at  2.2 GHz, equipped with 32\,GB of DDR3 1600\,MHz  RAM and connected  via InfiniBand Mellanox  SX6036 and FDR switches. For our  experiments, 
  a full rack constituted  of  52
  standard compute  nodes was used.  \textsc{Monch} is  slated to stay
  in  service  without  hardware  upgrade  for  the  duration  of  the
  Exa2Green  project,  allowing   an  identical  configuration  to  be
  recreated for future assessments of the baseline.\\

\item \textsc{Pilatus} is a dual-socket eight-core Intel Sandy
  Bridge EP based cluster composed of 42 compute nodes and used as Piz
  Daint pre-post  processing cluster at  CSCS.  The 2 login  nodes and
  the 42 computes nodes consists in 11 twin-pair Intel E5-Series DALCO
  r2264i4t 2U scalable compute modules. Each module contains 4 compute
  nodes based  on two Intel Xeon E5-2670 processors operating  at 2.6\,GHz 
  connected  via InfiniBand Mellanox SX6036 and FDR  switches (56
  Gb/s speed), and is equipped with 64GB of DDR3 1600\,MHz RAM. For our
  experiments, a  full rack constituted  of 42 standard  compute nodes
  was  used. \textsc{Pilatus}  is based  on  Intel's second-generation
  processors, conventional in  HPC systems and known to  be more power
  consuming than Ivy Bridge successor.\\

\item \textsc{Tintorrum} is  a heterogeneous cluster composed of
  28 compute nodes at UJI. For our  experiments only a subset of 16 nodes was considered,  
  Each of nodes of this subset includes a dual-socket hexa-core Intel Xeon E5645 processors 
  running at 2.40\,GHz, equipped with 288\,GB  of DDR3 1333\,MHz and connected via Infiniband QDR with a Mellanox MTS3600 switch. A  full tracing  experiment is  conducted on
  \textsc{Tintorrum} in order to  capture an overall  power profile at a much
  finer resolution. Also we analize the contribution of the MPI library
  to the energy consumption and how the use of the blocking versus polling
  message-passing policies can potentially render energy savings.
\end{itemize}

\subsection{Power measurement framework}
\label{subsec:3.3}

We present here the two different  frameworks deployed on the testbed HPC systems to
measure  performance and  power  consumption of the baseline execution of COSMO-ART

\subsubsection{E3METER metering framework}

Supercomputer  clusters considered  for our  experiments at CSCS of ETH Zurich are equipped with
E3METER Intelligent  Power Strips (IPS)  and Monitors (IPM)  which are
high   quality   electricity  meters   released   by  Riedo   Networks
\citep{Riedonetworks},   that  enable   to  monitor   and   log  power
consumption  of the IT  infrastructure as  well as  constantly analyze
line  voltage,  current, power-factor,  frequency  with 1\,\%  accuracy.
Using  reliable narrowband  powerline communication  (PLC) technology,
all  metering and  power  quality  data from  each  IPS are  centrally
collected  by the E3METER  Data Concentrator,  via the  existing power
cables thus  avoiding the need for  extra cabling.  This  data is made
available via  SNMP, HTTP, TELNET  through the built-in  Fast Ethernet
port.   Time  synchronisation  is  guaranteed by  using  NTP  servers.
Measured  data is  accessed  through the  open  source Cacti  software
including the E3METER Cacti Plugin  to scan the entire PLC network and
monitor in real-time the power usage of individual rack, recorded in 5
minutes interval periods.

\subsubsection{Power-performance measurement framework}

For our experiments at Jaume I University (UJI), we employed a version
of the integrated framework presented in~\cite{energy13} that employs the \pmlib library in
combination with Extrae and Paraver, which  are profiling/tracing and
visualization tools, respectively.

\textsc{COSMO-ART}  is compiled  using the  Extrae  compiler wrappers,
which automatically  instrument the Fortran code of  the model.  Next,
\textsc{COSMO-ART}  is  run on  the  nodes,  thus dissipating  certain
amount  of power.   These  nodes are  connected  to power  measurement
devices that account for the dissipated power/consumed energy and send
the power  data to the  tracing server.  The client,  meanwhile, sends
start/stop  primitives  in  order  to  gather  captured  data  by  the
watt-meters onto the  tracing server, where an instance  of the \pmlib
server is running.

Once  the run  is finished,  a file  containing the  power  profile is
created using the power data  received from the tracing server and the
instrumentation post-processing generates the performance trace files.
All  these files  are  combined then  in  Paraver which  allows us  to
visualise   the   performance  trace   and   the   power  profile   of
\textsc{Cosmo-art} all together.

@@

The \textsc{Cosmo-art}  baseline is a  pure MPI based  Fortran~90 code
currently  running  on   distributed  multi-core  systems  only.   The
software stack on the platforms was controlled using the modules
framework which gives an easy  and flexible mechanism to access to all
of the  provided compilers, tools  and applications.  For  our initial
benchmarking,   we  opted   for   the  GNU   compiler  gcc-4.8.1   on
\textsc{Monch} and gcc-4.8.2 on  \textsc{Pilatus} and \textsc{Tintorrum}) using the -O3 compiler
flag  in favor  of the  Intel compilers 14.0.1,  delivering inferior
performance.  In  addition, we installed the following implementations of MPI: MVAPICH2 1.9 on \textsc{Monch} and \textsc{Pilatus} and  OpenMPI 1.6.5 on \textsc{Tintorrum}. On the other hand, we  used HDF5 1.8.12 and NetCDF  (4.3.1) libraries for  the management  of extremely  large and
complex data collections.  All computes nodes have an operating system
based on GNU/Linux featuring ``2.6.32-358.11.1.el6.x86\_64'' kernel in
\textsc{Monch}     and      ``3.0.101-0.15-default''     kernel     in
\textsc{Pilatus}.

A snapshot of the code,  which includes at least conceptually, all the
information needed  to reproduce the  energy-to-solution benchmarks of
\textsc{COSMO-ART}, was produced and run  on a 1040 cores using 20~MPI
tasks  per node on  \textsc{Monch} and  on a  1344 cores  using 16~MPI
tasks per node on  \textsc{Pilatus}.  The calculated region was mapped
to the participating processors using a 2D-partitioning strategy.  The
distribution along the $x$ and $y$ coordinates was defined by setting:
$nprocx=40$  and $nprocy=26$  for \textsc{Monch}  and  $nprocx=28$ and
$nprocy=24$  as  $nprocx$  is   usually  kept  bigger  than  $nprocy$.
Besides,  as this  version doesn't  make use  of the  traditional GRIB
library, we  specified $nprocio=0$  for GRIB I/O.   Hyperthreading was
not considered in this study  as it previously revealed that it always
led to higher energy-to-solution.
